---
title: "Calculation of alpha diversity"
author: "`r params$author`"
date: "`r format(Sys.time(), '%d.%m.%y')`"
header-includes:
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
- \usepackage{flafter}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
output:
  pdf_document:
    latex_engine: lualatex
    fig_width: 7
    fig_height: 6
    fig_caption: yes
    df_print: kable
    extra_dependencies: ["float"]
    includes:
      in_header: header.tex
  word_document: default
  html_document:
    df_print: kable
    theme: cerulean
    highlight: haddock
    toc: yes
    toc_depth: 3
    toc_float: TRUE
    collapsed: FALSE
    smooth_scroll: TRUE
    code_fold: show
params:
  author:
    label: "author of the analysis"
    value: "no author given"
  workdir:
    label: "working directory"
    value: "U:/00_Shiny/Test"
  TU_type:
    label: "type of taxonomic units analyzed"
    value: "OTU"
  var_test:
    label: "x-axix variable for alpha diversity quick plot"
    value: NULL
  var_color:
    label: "color variable for alpha diversity quick plot"
    value: NULL
fontsize: 10pt
urlcolor: blue
papersize: a4
---
```{r version, echo = FALSE}
script_version <- "1.20"
#script_date <- file.info("C:\\MyFolder\\test.csv")$mtime
```
This is 'Calculation of alpha diversity' script version **`r script_version`** by Dr. Roman Gerlach.

## Load required libraries
```{r libraries, warning = FALSE, message = FALSE}
setProgress(0, message = "Loading require libraries...")
library(tidyverse)
library(phyloseq)
library(egg)
library(rmarkdown)
library(pander)
```
```{r global_options, echo = FALSE}
# Set up global options for nice reports and keeping figures:
knitr::opts_chunk$set(fig.width = 7, fig.align = "center", out.extra = "",
                      warning = FALSE, message = FALSE, fig.pos = "!H",
                      dev = "cairo_pdf",
                      cache.path = params$workdir)
options(knitr.kable.NA = '')
```
```{r general_options, echo = FALSE}
# setting TU type
TU_type <- ifelse(params$TU_type == "OTU", FALSE, TRUE)

```
```{r environment, echo = FALSE}

```
```{r alpha_plot_parameters, echo = FALSE}
var_test <- params$var_test
var_color <- params$var_color
```
## Setting up environment and script parameters
Type of imported taxonomic units (TUs)
`r ifelse(params$TU_type == "OTU", paste0("Taxonomic data is present in the form of **OTUs** (Operational Taxonomic Units)"), paste0("Taxonomic data is present in the form of **ZOTUs/ASVs** (Zero-radius Operational Taxonomic Units/Amplicon Sequence Variants)"))`

Define file paths:

* meta_file: **`r pander::pander(meta_file)`**

Set variables for quick alpha diversity plot (phyloseq function):

* var_test: **`r pander::pander(var_test)`**
* var_color: **`r pander::pander(var_color)`**

\newpage
## Calculate (phyloseq) alpha diversity
```{r calc_import_alpha}
setProgress(0.3, message = "Calculating alpha diversities...")
Sys.sleep(1)
set.seed(42)
# read list of Physeq objects to workspace
physeq_list <- readRDS(ifelse(TU_type, input$uploadZOTUphyseq$datapath, input$uploadOTUphyseq$datapath))

# Define Alpha diversity measures to be calculated: "Observed", "Chao1",
# "ACE", "Shannon", "Simpson", "InvSimpson", "Fisher"
meas <- c("Observed", "Chao1", "ACE", "Shannon", "Simpson", "InvSimpson")

# import metadata from physeq list (filter meta according to unfiltered PhyloSeq object is redundant?)
phyloseq_meta <- phyloseq::sample_data(physeq_list[["ori_no_ctrls"]]) #%>%
  # install speedyseq first!
  # rename_sample_data(sample_ID = sample)

# function to calculate Alpha diversity from phyloseq object
# merge with metadata
alpha_add_meta <- function(x, measure = meas) {
  # import and transform data
    # this is where alpha diversity is calculated
    ad <- phyloseq::estimate_richness(x, measures = measure) %>%
      tibble::rownames_to_column(var = "sample_ID") %>%
      dplyr::select(-starts_with("se")) %>%
      dplyr::left_join(phyloseq_meta, by = "sample_ID") %>%
      tidyr::pivot_longer(all_of(meas), names_to = "measure") %>%
      # remove NA --> check which samples!
      dplyr::filter(!is.na(value)) 

  # convert "NaN" to NA if present
  if(is.character(ad$value)) {
    ad <- dplyr::mutate(ad, value = na_if(value, "NaN"))
  }
  # remove unused factor levels
  ad <- ad %>%
    droplevels()
}
```
\newpage
## Plot alpha diversity
```{r alpha_plot, , echo = FALSE, fig.cap = "Different alpha diversity metrics before (A) and after (B) sample filtering"}
## calculate and plot alpha diversity from original and sample-filtered ('no out samples') phyloseq object
setProgress(0.6, message = "Quick plotting alpha diversities...")
alpha_plot_list <- setNames(vector("list", 2), c("original", "no_out"))
alpha_plot_list[["original"]] <- phyloseq::plot_richness(physeq_list[["ori_no_ctrls"]], x = var_test, color = var_color, measures = meas)
alpha_plot_list[["no_out"]] <- phyloseq::plot_richness(physeq_list[["no_ctrls"]], x = var_test, color = var_color, measures = meas)
alpha_plot_list <- lapply(alpha_plot_list,
            function(x) {x + egg::theme_article() + ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))})

# Combine each plot into one graphic.
alpha_plots <- egg::ggarrange(plots = alpha_plot_list, nrow = 2, labels = c("A", "B"),
               label.args = list(gp = grid::gpar(cex = 1.5, face = "bold")))
```
## Export alpha diversity data as list in RDS format
```{r export_phyloseq}
# generate empty list to store different alpha diversity measures
alpha_list <- setNames(vector("list", 2), c("original", "no_out"))

# calculate alpha diversity, combine with metadata and save to list
alpha_list[["original"]] <- alpha_add_meta(physeq_list[["ori_no_ctrls"]])
alpha_list[["no_out"]] <- alpha_add_meta(physeq_list[["filt_no_ctrls"]])

# save list of all alpha diversity data to file
setProgress(0.9, message = "Saving...")
saveRDS(alpha_list, file = paste0(
  output_dir2, "/alpha_list_", ifelse(TU_type, "zotu", "otu"), ".rds"))

```

\newpage
## Record session information
\tiny
```{r session_info, echo = FALSE}
# https://stackoverflow.com/questions/25646333/code-chunk-font-size-in-rmarkdown-with-knitr-and-latex
print(sessionInfo())
```
